Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-11 response-to-reviews.Rmd

*[issue #92](https://github.com/EcoClimLab/vertical-thermal-review/issues/92)*

**We have revised the first sentence to read: "Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry. We also choose to keep this rather obvious sentence because we feel that basic topic sentences are important for helping to guide readers from a diversity of sub-disciplines through the manuscript.**

**We separated VAZ section from VOC, and provided further context to NPQ in the following paragraph:**

- **Heat and associated VPD can indirectly impact the photosynthetic machinery by reducing the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition. Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species (ROS) [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a]. A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015]. This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry. Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007]. Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table 1, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a]. In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10˚C) during sunflecks [@waySunflecksTreesForests2012]. In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**

### 2022-02-11 vertical-thermal-review.Rmd
working on Tcrit paragraph

At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress]. 
Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021]. Thermal thresholds for photosystem II functioning serve as good proxies for temperatures at which leaves get necrotic [@krauseHightemperatureToleranceTropical2010].
Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]. 
Thermal tolerances vary globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
For example, *Quercus muehlenbergii* Engelm. growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx. growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. 
In an Australian savanna, crowns of *Acacia papyrocarpa* Benth. showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. 
Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019]. 
The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

"Thermal thresholds for leaves have also been hypothesized to vary with leaf temperature, and thus be higher in sun-exposed upper canopy leaves than in understory leaves. The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II (PSII) functioning using chlorophyll fluorescence techniques (Krause et al. 2012). Considering sun and shade leaf differences within canopy (Table 1), the temperature at which PSII efficiency declines, Tcrit, and at which it has decreased by 50%, T50, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama (Slot et al., 2019). In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019). These patterns of higher heat tolerance in the warmest leaves are consistent with interspecific patterns across temperature gradients (O’Sullivan et al. 2018; Feeley et al., 2020; Slot et al., 2021a), and experienced maximum temperatures (Perez & Feeley, 2020). However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences [sorry, bit of a blabla sentence to wrap up]."

>>>>

Thermal thresholds for leaves have also been hypothesized to vary with leaf temperature, and thus be higher in sun-exposed upper canopy leaves than in understory leaves. The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II (PSII) functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. Considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), the temperature at which PSII efficiency declines, $T_{crit}$, and at which it has decreased by 50%, $T_{50}$, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019]. At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress]. 
Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021]. Thermal thresholds for photosystem II functioning serve as good proxies for temperatures at which leaves get necrotic [@krauseHightemperatureToleranceTropical2010].
Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]. 
Thermal tolerances vary globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
For example, *Quercus muehlenbergii* Engelm. growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx. growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. 
In an Australian savanna, crowns of *Acacia papyrocarpa* Benth. showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. 
Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019]. 
The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

 "In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019). These patterns of higher heat tolerance in the warmest leaves are consistent with interspecific patterns across temperature gradients (O’Sullivan et al. 2018; Feeley et al., 2020; Slot et al., 2021a), and experienced maximum temperatures (Perez & Feeley, 2020). However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences [sorry, bit of a blabla sentence to wrap up]."
